\Large \bf Cyclone: Warp Speed Replication for Key Value Stores

### 2017-04-19 paper/paper.tex
Updates to paper.

\documentclass[10pt, preprint]{sigplanconf}
\usepackage{times,epsfig,endnotes,graphicx}
\usepackage{datetime, url, hyperref, pdfpages}

@@

\begin{abstract}
Persistent key value stores are rapidly becoming an important component of many
distributed data serving solutions - with innovations targeted at taking
advantage of growing flash speeds. Unfortunately their performance is hampered
by the need to maintain and replicate a write ahead log to guarantee
availability in the face of machine and storage failures. Cyclone offers a
solution to this problem by leveraging a small amount of directly attached
non-volatile memory such as NVDIMMs to maintain a two level log - the upper
level in a small amount of NVDIMM memory draining into a larger lower level log
maintained on flash. Replicating the smaller log kept in NVDIMMs transforms the
replication problem into a packet switching problem - the leader sends the exact
same packet to all follower replicas and we solve the problem as such by using
the RAFT protocol implemented as a software multicast module on the Data Plane
Development Kit (DPDK). Second, we levergage the observation that key value store
interfaces are commutative for operations to different keys to scale our two
level log horizontally into a number of physical two level logs - with a hash of
the key being used to choose the right log. Each log is replicated by an
independent instance of RAFT, thereby leveraging the scalability afforded by
classic software packet switching when using multiple CPU cores. Cyclone is able
to replicate millions of small updates a second using only commodity 10 gigabit
ethernet adapters and improves the performance of RocksDB - a popular persistent
key value store used in industry by orders of magnitude when compared to its own
write ahead log without replication.
\end{abstract}

\section{Introduction}
Memory technology in datacenters is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types such
as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
programmers to build durable data structures on non-volatile heaps have become
available~\cite{mnemosyne, nvheaps, cdds} - an example of such a library that we
have used in this paper is Intel's non volatile memory library
(NVML)~\cite{nvml}. A missing piece however is a component to make such durable
data structures highly available using replication across a commodity network
such as ethernet. Existing solutions that offer good performance are dependent
on high performance RDMA networking hardware, with a dependence on external
services (such as zookeeper~\cite{zookeeper}) and application specific code for
fault recovery~\cite{farm, htm}.  This is a problem for programmers using NVML
where one cannot apriori guarantee the availability of such RDMA networking
hardware (for e.g., when running on externally hosted environments or public
clouds).

Cyclone is prototype replication middleware aiming to demonstrate that one
can add fault tolerance transparently to a specific class of client server
applications written on top of libraries such as NVML. We specifically target
key-value maps in this paper and in that context make two key
contributions. First, we propose a simple API that enables server state to be
replicated without the programmer needing to worry about fault recovery. This
allows one to start by building a concurrent durable data structure using
libraries such as NVML and add availability at a later stage using Cyclone. The
second contribution is to show that one can achieve reasonably good replication 
performance using commodity ethernet via innovations in the networking stack and
by scaling replication with the application. This result should be a useful
design point for developers looking to deploy applications on non-volatile heaps
in commodity ethernet-based environments such as public clouds.

\section{Single Node Programming Model}
Cyclone assumes a concurrent durable application on a non-volatile heap as the
starting point. A number of libraries and filesystems make this possible
today~\cite{nvml, dax, pmfs, mnemosyne, nvheaps, cdds}. Cyclone currently
supports applications built on top of Intel's Non Volatile Memory Library
(NVML~\cite{nvml}). We describe NVML's programming model in this section to
provide adequate background for the rest of this paper. NVML requires 
that code that makes changes to the non-volatile heap is enclosed in a crash
consistent transaction such that either all of the transaction executes or none
of it executes in the event of power failures.
%This is critical to maintaining consistency as the data structure, being
%durable, must survive across power failures.
Figure~\ref{fig:example} illustrates the use of NVML to build a persistent
linked list. Updates to the linked list are enclosed in a NVML transaction to
avoid corruption due to power failure. It is also worth noting that NVM
libraries such as NVML usually allow remapping non-volatile heaps to a different
virtual memory address range across restarts (e.g., using {\tt TOID} macros to
implement fat pointers in NVML).

\begin{figure}
  { \scriptsize
\begin{verbatim}
struct ll_node {
  int value;
  TOID(ll_node) next;
};

void insert_after(TOID(ll_node) prev, 
                  TOID(ll_node) new_node)
{
  TX_BEGIN {
    TX_ADD(new_node);
    D_RW(new_node)->next = D_RO(prev)->next;
    TX_ADD(prev);
    D_RW(prev)->next = new_node;
  } TX_END
}

\end{verbatim}
  }
\vspace{-0.2in}
\caption{A Persistent Linked List}
\label{fig:example}
\end{figure}

It is easy to turn the linked list example in NVML into a concurrent linked list
by simply adding a lock to it. Such a lock can be maintined in volatile DRAM if
so desired for performance. More sophisticated fine-grained or optimistic
concurrency control mechanisms for NVM are also
possible~\cite{mnemosyne, cdds, nvheaps}.

\section{Replication}
Cyclone is designed to add fault tolerance to NVML client server applications
via replication -- a quorum of machines maintains \emph{equivalent} state.
Server state is queried and manipulated via RPC calls from clients.  Cyclone
treats the RPC call itself as a variable sized opaque blob of data, leaving the
application free to choose how it wishes to marshall call related details and
arguments.

Cyclone replicates the RPC call across replicas rather than replicating
every access (to the server state) made during execution of the RPC call.
Cyclone provides strongly consistent replication - every machine in the quorum
of replicas maintains a log of RPC calls.  All machines agree on the committed
sequence of RPC calls in the log. We achieve this by running an instance of the
RAFT~\cite{raft} consensus protocol to keep the log of RPC calls on different
machines in synchronization. Specially flagged `read-only` RPC calls are
executed only on the leader replica and are not replicated.

\subsection{Decoupled Replication}
\label{sec:decouple}
The standard replicated state machine approach requires that commands (entries
in the log) be committed before they can be applied to the state machine --
i.e. in our case before the RPC call can be executed. This is because
replication of an accepted log entry can fail for a
number of reasons such as change of leadership and subsequent
rollback of previously appended (but not committed) log entries.

In contrast to other systems, Cyclone decouples execution of an RPC call on any
replica from its replication by executing the RPC call as soon as it is received
on any replica. We exploit the fact that RPC calls that modify state must
themselves be executed in a crash consistent transaction. Cyclone delays the
actual commit of the crash consistent transaction until after the replication is
complete. In the event that replication fails, the transaction is aborted and
any changes to the non volatile heap are rolled back. Figure~\ref{fig:async_rep}
shows a pseudocode description of how the runtime handles \emph{decoupled
  replication}.

\begin{figure}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments 
TX_BEGIN { 
  Execute RPC call and modify local NVM 
  Block till replication result known 
  if(replication failed) {
     TX_ABORT 
  } 
} TX_END
\end{verbatim}
}
\caption{Decoupled Replication}
\label{fig:async_rep}
\end{figure}

Decoupled replication provides good performance and we believe it would be the
common case for RPC calls replicated via Cyclone. However the program might on
occasion require successful replication to be a precondition to exection. A case
where we find this necessary is \textit{output commit}~\cite{output_commit} -
where the RPC call has external side effects (e.g., communication with other
machines in the cluster) and therefore crash consistency is insufficient to roll
back the execution of the RPC call in the event that replication fails. We
therefore provide users the ability to declare that an RPC call should execute
in a coupled manner. Cyclone would then ensure that the RPC call is successfully
replicated before it attempts execution on any of the replicas.

\subsection{Network Stack}
Cyclone is designed around DPDK~\cite{dpdk} -- a high performance network stack
optimized for Intel processors. DPDK provides convenient access to packets
arriving at an ethernet NIC by moving them into packet buffers and directly
placing those buffers into the last level cache of the CPU during DMA~\cite{ddio}.

Cyclone separates the available CPU cores on a machine, in a configurable way,
into cores dedicated to work related to the (RAFT) consensus protocol and cores
decidated to running application code. Each RAFT core runs an \emph{independent}
instance of the RAFT protocol - in conjunction with a similar instance on every
other replica. RAFT cores pass work to the application cores through lock-free
double ended queues. A RAFT core dispatches an incoming RPC call for execution
simultaneously with sending it out for replication.

Each RAFT core has a dedicated NIC queue pair. The input queue receives RPC
calls from clients as well as RAFT protocol related traffic. The output queue
only carries RAFT protocol related traffic. Each application core has a
dedicated \emph{output} queue on the NIC. When completed, the response to the
RPC call is sent from the application core via its dedicated output queue. This
division of NIC queues eliminates the need for synchronization between RAFT
cores and application cores when accessing the NIC.

Cyclone can exploit multiple NICs on a machine by binding each RAFT or
application core to a specific NIC and picking the necessary queues from that
NIC.

\subsection{Zero Copy Batching}
Consensus protcols such as RAFT usually have a simple common case path. Once a
quorum leader is elected, the leader's primary job is to transmit new entries
appended to its persistent log to its followers. The primary job of a follower
is to receive these entries, append them to its persistent log and send
acknowledgments to the leader. An entry is considered committed when the leader
receives an acknowledgment for an entry from a majority of the quorum. Followers
receive intimation of committed entries as piggybacked information on subsequent
replication messages. The common case replication path at the leader lends
itself to \emph{zero copy} transmission to the replicas, since the same set of
log entries is sent to all replicas.

In the receive path, we avoid an additional copy of the incoming RPC call
contents from DRAM to persistent memory by allocating space for packet buffers
on the non-volatile heap. However, since packet data is cached, Cyclone must
ensure durability (or persistence) of this data by explicitly flushing the dirty
cachelines~\cite{pmfs}. Cyclone maintains its log as a circular sequence of
pointers to packet buffers and therefore the append operation is completed by
adding a pointer to the received packet and flushing the cacheline containing
the pointer.

We exploit the fact that DPDK allows prepending data to a packet
without physically moving the data in a packet buffer. The received packet
contains only the RPC call from the client. We prepend a RAFT related header
(containing the previous log index, previous log term, leader's term and
leader's commit point~\cite{raft}) to the packet buffer itself. To this RAFT header 
we then prepend an IPv4 header. We assign pre-determined values to the
IP address fields in order to exploit 5-tuple steering at the receiving NIC for 
the placement of the incoming packet in the correct queue.
This composite packet buffer is then sent \emph{unchanged} to every replica.
To properly route the packet through switches, for each follower replica we
allocate a small buffer containing an ethernet header with appropriate addresses
and \emph{chain} the buffer in front of the prepared packet - thereby ensuring
that we do not need to copy the contents of the packet itself.

Although the common case path in RAFT is straightforward, it still requires
hundreds of cycles (or a few microseconds) to process each log entry, which is a
significant amount in the face of sub 10 us network latency between nodes with
DPDK. We amortize this overhead when operating under load by batching multiple
RPC calls together and treating the batch as a single RAFT log entry.  DPDK by
itself can efficiently batch incoming packets under load, therefore allowing the
user to easily receive a burst of packets (currently limited to 32) rather than
a single packet for a polling call.  DPDK further allows chaining packet buffers
together and treating the chain as a single packet. We chain these packet
buffers together to form a single packet. Since the entire chain is sent as a
single packet when being replicated, the length of a chain is limited by the
size of the coalesced packet, that must fit into an ethernet jumbo frame of
approximately 9KB. Figure~\ref{fig:zc_batch} summarizes this set of
transformations to incoming request packets from clients at the leader replica.

\begin{figure}
\includegraphics[width=0.9\columnwidth]{figures/chain}
\vspace{-0.15in}
\caption{Zero Copy Batching}
\label{fig:zc_batch}
\end{figure}

\subsection{RPC Call Execution}
\label{sec:exec}
We now describe how RPC calls are executed and results returned to the
client. There are two important aspects of RPC call execution in Cyclone -
steering and at most once semantics.

An RPC call via Cyclone must traverse a specific RAFT core (and associated
consensus protocol instance) and specific application core. These details are
provided by the user when making the RPC call to the Cyclone client side
library. The user therefore is aware of the number of execution cores and RAFT
instances and given complete control of \emph{RPC call steering}. For most
applications we envisage a simple modulo hashing scheme should suffice for load
balancing, something that we plan to later automate in Cyclone.

Cyclone provides at most once semantics. It supports a large but fixed (at
startup time) number of clients. Each RPC call from a client bears the client
number and a sequence number. \emph{Each execution core independently enforces
  that accepted RPC calls from the same client bear increasing sequence
  numbers}. This is done by storing the last executed sequence number for every
client in a per-core durable data structure as part of the crash consistent
transaction wrapping the execution of the RPC call. Even in the presence of
fail-overs, this durable memory of sequence numbers seen from a client ensures
at most once semantics. In addition to guarding against repeated execution of
the same call, Cyclone also remembers the \emph{result} of the last executed RPC
call from every client. This is done to assist clients - particularly stateless
clients. Cyclone allows the client to query the last seen sequence number and
result for that RPC call from any application core on the quorum of servers.

\section{A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practice by putting together an
example of a linearizable hash table - a useful and common data structure.
The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys.

A durable hash table is simple to derive from a volatile hash table and readily
available in NVML as an example~\cite{nvml_hash}. To use that (non
client-server) example with Cyclone we added a simple RPC call format, manually
marshalled and demarshalled the few arguments to the RPC call in an RPC call
handler - a few tens of lines of code. Finally, we added a simple client driver
program (a few tens of additional lines of code).

In order to use the hash table with Cyclone, we simply need to to declare the
RPC call handler as the point of entry to Cyclone at the server end.  The client
end issues RPC calls via Cyclone's client side library. Crucially neither side
involves fault awareness or fault recovery code - a key goal for Cyclone.

There are two interesting challenges in scaling the performance and improving
the utility of the linearizable and durable hash table. The first is to ensure
linearizability while scaling performance by adding RAFT cores and application
cores. The second is to support distributed transactions - that manipulate a set
of key-value pairs atomically in the face of failure.

\subsection{Scaling}
Scaling the performance of the hash table by using multiple RAFT cores and
application cores presents determinism related challenges. Two RPC calls being
steered through different RAFT cores can end up executing in different orders on
different replicas even if steered to the same application core. At the same time,
two RPC calls steered through the same RAFT core but dispatched to different
application cores can end up completing in different orders on different replicas.

Our solution to this problem in Cyclone is to observe that replicas of the hash
table require \emph{per-key determinism} and not determinism across \emph{all}
operations. We therefore steer operations to a particular key to the same RAFT
core and the same application core (using a hash on the key).  The actual
configuration of the hash table therefore can be different across different
replicas - for example due to the insertion of two keys mapping to the same hash
bucket completing in different orders across different replicas. However, since
the results of insert, delete and lookup operations for a key depend only on
previous operations to the same key the replicas provide identical results for
committed operations. \emph{In other words, we exploit the fact that operations to
different keys in a map are commutative.}

The RPC call steering mechanism described above leaves the programmer free to
use any synchronization mechanism they wish between concurrent accesses from the
application cores to the data structure, since the ordering between accesses
from different cores is allowed to be non-deterministic. For our open addressed
hash table we chose the simple scheme of a per-bucket lock.

\subsection{Distributed Transactions}
A useful primitive for many distributed systems is the ability to run
distributed transactions - an atomic sequence of conditional updates spanning
multiple groups of replicas. Distributed transactions have two distinct facets:
atomicity and fault tolerance. Distributed transactions appear atomic with
respect to other atomic transactions and simple updates or reads by virtue of
standard concurrency control techniques. A common technique is strict two phase
locking: acquire all locks, update the locked objects and finally release all
the locks. We believe that NVM application programmers - particularly those
with experience in concurrency should not find it hard to deal with
distributed concurrency. The hard part for most programmers is likely to be the
second aspect: fault tolerance. A distributed transaction spans many machines
and has a large footprint to clean up in the event of failure. Failure handling
is made more complex when considering cascading failures - further failures
while recovering from a failure. Cyclone was designed to allow primitives such
as distributed transctions to be added without the programmer needing to worry
about faults or writing fault recovery code. As an exercise we show how to add
support for distributed transactions to our hash table.

We support distributed transactions in our hash table by allowing the user to
specify a group of insert/delete/lookup operations that must occur
atomically. We use the existing bucket locks to support a simple two phase
locking protocol to implement the transction. A client makes an RPC call to the
server specifying the transaction and the server executes it on the client's
behalf. Execution of the transaction has four distinct phases: acquire locks,
verify values, apply updates and finally release locks.

We guard against faults by executing the the entire transaction on a quorum of
machines via Cyclone. All machines in the quorum execute the transaction, making
calls to other quorums to execute steps in the two phase locking protocol. Since
execution of the RPC call leaves \emph{externally visible side effects} by
virtue of the further RPC calls it makes, we require output
commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
complete once it starts. Hence, for distributed transactions, we turn off
decoupled execution (Section~\ref{sec:decouple}).

Next, we need to ensure that although every machine in the quorum makes all
necessary RPC calls, the call itself is executed only once. This is achieved by
using Cyclone's at most once semantics. All machines in the quorum executing the
transaction use the same client identifier and the same montonically increasing
sequence of RPC call numbers. Finally, we need to deal with the fact that
Cyclone only remembers the last executed RPC call for a client. This means that
if different machines in the quorum move at different speeds, some machines will
not receive a response to the RPC call - they will instead receive an error that
tells them that they are too far behind.

To ensure that the transaction is still guaranteed to complete, we use a
\emph{different} client identifier for each \emph{phase}. Thus, until the
transaction completes and a new one is started the response for any RPC call
continues to be available. Figure~\ref{fig:dist_tx} shows in pseudocode the
server side details of a transaction. Once initiated, the distributed transaction
is guaranteed to complete given availability of a quorum.
%However, if the client
%requires intimation of completion, the final transaction state can be written to
%a distinguished partition as part of the transaction itself.

Distributed transactions built around at most once semantics are not intended
for performance. On the other hand, at most once semantics allows users of
Cyclone to avoid writing explicit fault recovery code as compared to fault
recovery in systems that explicity require such cleanup (~\cite{farm}, pp
59--63). Under the assumption that distributed transactions will be infrequent
compared to single key operations, such simplicity is key to increasing the
adoption of NVM among datacenter operators concerned about availability.

\begin{figure}
{ \scriptsize
\begin{verbatim}
wait until RPC call is replicated
respond to client saying tx accepted for execution
acquire:
 set client identifier to 1
 for each machine
   Acquire write or read locks on the machine
   If machine reponds RPC sequence too old
     Transaction completed by someone else

update:
 set client identifier to 2
 for each machine
   Update object
   If machine reponds RPC sequence too old
     Transaction completed by someone else

release:
 set client identifier to 3
 for each machine
   Update object
   If machine reponds RPC sequence too old
     Transaction completed by someone else
\end{verbatim}
}
\vspace{-0.22in}
\caption{Distributed Transactions}
\label{fig:dist_tx}
\end{figure}

@@

\section{Conclusion}
Cyclone allows developers to add availability to key value maps built on top of
libraries such as NVML and achieve good performance on commodity networking
hardware. A key question is whether Cyclone can be used for other applications
such as filesystems? \emph{We believe that the requirement for scaling across
  multiple replication quorums can be satisfied by applications where the
  scalable commutativity rule~\cite{scalable_commutativity} applies to common
  case primitives.} Although our intitial release of Cyclone will be for key
value style applications, our on-going research is based on the scalable
commutativity hypothesis.
\newcommand\myurl[2]{\url{#1}}
\bibliographystyle{plainnat}
\bibliography{paper}

>>>>

\documentclass[10pt, preprint, nonatbib]{sigplanconf}
\usepackage{times,epsfig,endnotes,graphicx, verbatim}
\usepackage{datetime, url, hyperref, pdfpages}

@@

\begin{abstract}
Persistent key value stores are rapidly becoming an important component of many
distributed data serving solutions - with innovations targeted at taking
advantage of growing flash speeds. Unfortunately their performance is hampered
by the need to maintain and replicate a write ahead log to guarantee
availability in the face of machine and storage failures. Cyclone offers a
solution to this problem by leveraging a small amount of directly attached
non-volatile memory (NVM) such as NVDIMMs to maintain a two level log - the upper
level in a small amount of NVM draining into a larger lower level log
maintained on flash. Replicating the smaller log kept in NVMs transforms the
replication problem into a packet switching problem - the leader sends the exact
same packet to all follower replicas and we solve the problem as such by using
the RAFT protocol implemented as a software multicast module on the Data Plane
Development Kit (DPDK). Second, we levergage the observation that key value store
interfaces are commutative for operations to different keys to scale our two
level log horizontally into a number of physical two level logs - with a hash of
the key being used to choose the right log. Each log is replicated by an
independent instance of RAFT, thereby leveraging the scalability afforded by
classic software packet switching when using multiple CPU cores. Cyclone is able
to replicate millions of small updates a second using only commodity 10 gigabit
ethernet adapters and improves the performance of RocksDB - a popular persistent
key value store by orders of magnitude when compared to its own write ahead log
without replication.
\end{abstract}

\section{Introduction}

\section{Two Level Log}
A fundamental challenge in building an efficient replicated log is dealing with
network overheads. Cyclone'e construction is predicated on the observation that
in the common case protocols such as RAFT simply multicast the same entry
received from a client, from the leader replica to all follower replicas. This
suggests that an efficient way to implement a replicated log is to treat it as a
software packet switching problem. However peristing logs on flash represents an
impediment in that block storage devices are not good candidates to hold packets
being software switched. Our solution is to split the log into two levels. The
upper level is held in persistent directly attached non-volatile memory (NVM)
such as NVDIMMs. The lower level is held on a flash SSD - we term the lower
level log as the flash log. We periodically drain entries in batches from the
NVM log to the backing flashlog. An important and beneficial side effect of this
two level approach is that we need only limited amounts of NVM with the bulk of
the log residing on flash - an arrangement that maps well to the relative cost
of the two types of memory. We begin by describing the NVM log and how we
implement the RAFT consensus protocol in the packet switching paradigm around
it. We then describe the flashlog and how we drain entries in batches from the
NVM level to the flash level.

\subsection{NVM Log}
The NVM log is maintained as a circular log of fixed sized pointers (a pointer
log) to memory buffers - each of which holds a client request. This arrangement
is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of
such an arragement. Adding a level of indirection in the NVM log allows us to
isolate the circular log being manipulated by the RAFT protocol from the
problems of variable sized client requests that can be scattered across
non-volatile memory depending on the internals of the memory allocator of the
packets switching library we use. Another advantage is that it makes recovery
from NVM easy - appends to the circular pointer log are atomic and we use the
pointer log to recover allocator state i.e. what pieces of NVM are currently in
use by the log.

\begin{figure}
\centering
\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
\hspace{-0.2in}
\caption{NVM Log Structure}
\label{fig:nvm_log}
\end{figure}

We now describe how we implement the RAFT consensus protocol to replicate the
NVM log across a set of replicas. The software packet switching idea we apply is
to separate the control plane from the dataplane using the Data Plane
Development Kit (DPDK~\cite{dpdk}) library for direct userspace access to the
network interface.

\subsubsection{Dataplane Separation}
We use RAFT~\cite{raft} as the underlying protocol for log replication. RAFT
replicates a log across a set of replicas. Its operations are leader-based and
RAFT incorporates a leader election algorithm that elects a leader with the most
upto date log. The leader receives log entries from clients, appends them to the
persistent log (NVM log in this case) and multicasts the new entry to follower
replicas. On a response from enough follower replicas (constituting a majority
quorum) the leader replica can then execute the command in the log entry. In
our case this involves accessing the key value store and sending the response
back to the client.

We implement the common case flow in RAFT as a software packet switch using
DPDK. The dataplane now encompasses the NIC and the CPU caches with the NIC
placing packet data directly in the CPU cache via DDIO~\cite{ddio} - a standard
feature of most NICs today. The dataplane for the RAFT protocol is therefore
separated from the control plane. The control plane now runs on a CPU core and
takes care of checking adherence to the protocol: checking for correct term, log
index and that the replica receiving the packet from the client is indeed the
leader. In addition, the control plane is responsible for non common-case
activities such as dealing with timeout and leader elections.

The pseudocode in Figure~\ref{fig:control_plane} describes the control plane in
Cyclone organized as event handlers triggered on receiving a packet. We focus on
two events for illustration (and brevity): the event where a request is received
from a client and the event where a leader election related message is
received. A key concern when we built Cyclone was adhering as far as possible to
the software packet switching principle of separating the control plane from the
dataplane. This means that the CPU should avoid examining every byte of the
packet as far as possible.

If an election message is received all the code related to handling
the election related message is run on the CPU core and all contents of the
packet are examined to decide what to do in accordance with the RAFT protocol
state machine. This is an exception to the software packet switching principle
but one that we found acceptable since leader elections are rare compared to the
common case of replicating client packets.

The sequence of operations when a packet is received from the client however
needs careful design for efficient software packet switching. To illustrate how
this is achieved, Figure~\ref{fig:packet_layout} shows how Cyclone manipulates
packet layouts across the two steps of prepending a RAFT header and transmitting
to follower replicas. DPDK describes packets using an ``mbuf'' data
structure. Roughly speaking, an ``mbuf'' consists of a flat array of bytes
actually containing the byte and a fixed size piece of external metadata that
describes various aspects of the packet, most crucially a pointer to the start
and end of the packet in the byte array. DPDK's userspace drivers receive
packets from the NIC such that they are offset in the byte array by a
configurable amount referred to as ``headroom''. We prepend RAFT related
information in the headroom and shift the start pointer appropriately - these
operations are standard enough for software packet switches that DPDK provides
convenient library calls to do it. For the final step, we need to prepare the
packet for transmission to the various follower replicas. To do this we prepare
a different packet containing an ethernet header for the targeted replica and
``chain'' the data packet to each of these headers. Each header is then
separately handed off to the driver for transmission via the NIC.

\begin{figure}
\begin{verbatim}
    event_receive_client_req()
    {
      if(!check_is_leader()) {
        drop_msg
        return
      }
      Prepend raft log term and index
      Persist to local log
      Transmit to follower replicas
    }

    event_receive_election_msg()
    {
      Handle_election_msg()
    }
\end{verbatim}
\caption{Cyclone Control Plane Fragment}
\label{fig:control_plane}
\end{figure}

\begin{figure}
  \caption{Cyclone Network Packet Layout}
  \label{fig:packet_layout}
\end{figure}

TBD: talk about how we do not need a reliable network layer like TCP/IP.
\subsubsection{Batching}

\subsection{Flash Log}
TBD: talk about why we don't need group commit.
\section{Horizontal Scaling}

\subsection{API Commutativity}

\subsection{Core Assignment}

\subsection{NIC Queue Assignmnt}

\subsection{Ganged Operations}

@@

\section{Related Work}
Memory technology in datacenters is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types such
as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
programmers to build durable data structures on non-volatile heaps have become
available~\cite{mnemosyne, nvheaps, cdds} - an example of such a library that we
have used in this paper is Intel's non volatile memory library
(NVML)~\cite{nvml}. A missing piece however is a component to make such durable
data structures highly available using replication across a commodity network
such as ethernet. Existing solutions that offer good performance are dependent
on high performance RDMA networking hardware, with a dependence on external
services (such as zookeeper~\cite{zookeeper}) and application specific code for
fault recovery~\cite{farm, htm}.  This is a problem for programmers using NVML
where one cannot apriori guarantee the availability of such RDMA networking
hardware (for e.g., when running on externally hosted environments or public
clouds).

\section{Conclusion}
Cyclone allows developers to add availability to key value maps built on top of
libraries such as NVML and achieve good performance on commodity networking
hardware. A key question is whether Cyclone can be used for other applications
such as filesystems? \emph{We believe that the requirement for scaling across
  multiple replication quorums can be satisfied by applications where the
  scalable commutativity rule~\cite{scalable_commutativity} applies to common
  case primitives.} Although our intitial release of Cyclone will be for key
value style applications, our on-going research is based on the scalable
commutativity hypothesis.
\newcommand\myurl[2]{\url{#1}}
\bibliographystyle{plain}
\bibliography{paper}
